### 2021-02-26 rad-icfp-2021/efficientdiff.tex
adding some structure to the document
part of the operational semantics target
outline paper in intro
writing about target language

\section{Efficient reverse-mode}

@@

>>>>

\section{Efficient reverse-mode}
\label{sec:efficientrad}

@@

The last language we present, which we call the target language, is an extension to the source language. 
The types and terms are presented in Figure~\ref{fig:target_grammar}.
The language contains lambda abstractions for dealing with continuations. 
We are not interested here with partial applications and our language is pure, so functions take $n$ arguments.
Lambda abstractions and applications are going to be removed during partial evaluation and this notation makes reading slightly easier.

\input{figures/target_grammar}

In addition, we require the set of operations to be closed under partial differentiation. 
In more detail, for every unary scalar operation !op1!, 
we assumed given an operator $\partial_1$!op1! whose semantics should be the derivative of !op1!.
For every binary operator !op2!, we assume given operators $\partial_1$!op2!, $\partial_2$!op2! 
representing respectively the first and second partial derivative of !op2!.
Similarly, we need to add some functions on arrays. 
!scanl! is a !foldl! that stores all intermediate results. 
!scanr! is similar, but reads the array from right to left, and also stores from right to left.
!reduce! is a !foldl! for which the function is associative. 
This means in practice that it can be computed faster.
We also include a !map3! operator. 
Though not necessary as it can be simulated with two !map2!, it is a convenient notation and can somtimes run faster in practice.
Finally, we add a somewhat strange operator which we call !shift1!. 
It takes an array of size $n$,  shifts all the element of the array by one, 
forgetting the last element, and puts a one in the first place. 
This operator naturally shows up when differentiating !foldl!.

The type system for the extended grammar of target is in Figure~\ref{fig:target_typesystem}.

\input{figures/target_type_system}

A standard semantics of this language is given in Section~\ref{sec:correctness}.

We now present our transformation from target UNF to target. 
The transformation is presented in Figure~\ref{fig:unf_to_target}.

\input{figures/unf_to_target}

\begin{remark}
    One could rightfully wonder: where did reverse-mode really happened here?
    In a way, source UNF does not seem to do much. 
    Then $\Dsynrevsymbol$ is very simple on UNF.
    So perhaps everything happens when going from target UNF to target?
    \MH{finish or remove.}
\end{remark}

### 2021-02-26 rad-icfp-2021/simplediff.tex
writing about target language

Following the intuition higlighted in Section~\ref{subsec:insights}, we present a new language which we call unary form (UNF). 
It simply consists of a composition of unary operators. That said, we want to compile our source language to this intermediate representation, 
and we need to remember some information about the initial term.

@@

There are a few notable things in this syntax. 
Every term is indexed by a well-defined contex $\Gamma$ from the source language.
In addition, a variable has a integer index $1\leq i\leq n$ where $n$ is the size of the context $\Gamma$.
Every constant, unary, or binary operator is summarised as an $n$-ary operator !op!$_n$.
Sequential composition is denoted by !;! and !e1;e2! means that !e1! should be performed, and then !e2!.
Every array operator !map!, !map2!, !foldl! has an extra index which represents a well-formed term in the source language.

@@

We are now able to present a simple transformation for purely functional reverse-mode on UNF.
This transformation is somehow straightforward as some of the work was put into the design of UNF.
Our main takeaway for this paper is that reverse-mode is complicated beause there is a lot happening in the same time. 
But by decoupling the problems, it becomes muche easier. UNF is somehow dealing with the inherent type-dependency of reverse-mode.
Then the macro $\Dsynrevsymbol$ adds Jacobians. It allows to compute gradients in a compositional way.
Finally, we'll see in the next section how to compile back to a normal language and optimise our transformation to get efficiency.

The differentiation macro is given in Figure~\ref{fig:diff_macro}.

>>>>

Following the intuition highlighted in Section~\ref{subsec:insights}, we present a new language which we call unary form (UNF). 
It simply consists of a composition of unary operators. That said, we want to compile our source language to this intermediate representation, 
and we need to remember some information about the initial term.

@@

There are a few notable things in this syntax. 
Every term is indexed by a well-defined contex $\Gamma$ from the source language.
In addition, a variable has a integer index $1\leq i\leq n$ where $n$ is the size of the context $\Gamma$.
Every constant, unary, or binary operator is summarised as an $n$-ary operator !op!$_n$.
Sequential composition is denoted by !;! and !e1;e2! means that !e1! should be performed, and then !e2!.
Every array operator !map!, !map2! and !foldl! has an extra index which represents a well-formed term in the source language.

@@

We are now able to present a simple transformation for purely functional reverse-mode on UNF.
This transformation is somehow straightforward as some of the work was put into the design of UNF.
Our main takeaway for this paper is that reverse-mode is complicated because there is a lot happening at the same time. 
But by decoupling the problems, it becomes much easier. UNF is somehow dealing with the inherent type-dependency of reverse-mode.
Then the macro $\Dsynrevsymbol$ adds Jacobians. It allows to compute gradients in a compositional way.
Finally, we'll see in the next section how to compile back to a more standard language and optimize our transformation to get efficiency.

The differentiation macro is given in Figure~\ref{fig:diff_macro}.
One might be surprised that the macro does not seem to do much, and it is partly correct.
We tried to decompose reverse-mode into more elementary small steps, and each step is slightly non trivial and dealing with one aspect of reverse-mode.
The perhaps slightly more involved transformation is the one coming next,  going from target UNF to our target language.
